---
title: "KSPlots"
author: "Daniel Kick"
date: "12/9/2022"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tictoc)
# for phylogenetic trees
library(ape)
library(tidytree)
# used for computing HDI on distribution of differences
library(bayestestR)
# for easy italics in ggplots
library(mdthemes) 
theme_set(mdthemes::md_theme_minimal())

show_demo_plots = FALSE

# Shawn's requested colors:

#           RStudio     Hex
# ingroup	  "darkgreen"	#006400
# outgroup	"purple"	  #A020F0

# if not in or outgroup use black

```



# Load in data
Load in KS
```{r}
M <- read.table("./inst/extdata/fullks-subtribe.txt")
names(M) <- c("Species", "KS", "Group", "Subtribe")
# Subtribe == NA are in the Group == outgroup
M[is.na(M$Subtribe), c("Subtribe")] <- "outgroup"

# ks less than 3
M <- M[M$KS <= 3, ]
```


Load phylogenetic tree
```{r}
apeTree <- ape::read.tree(file = './inst/extdata/apro.tre')
# plot(apeTree)

tree_df <- as_tibble(apeTree)

# tree_df
```


Standardize naming
```{r}
# duplicate to ensure that the tree has the correct matches properly

M[M$Species == 'Sinapis_alba-BMAP', 'Species'] <-   'Sinapis_alba_BMAP'
M[M$Species == 'Eruca_vesicaria-BMAP', 'Species'] <-   'Eruca_vesicaria_BMAP'

tree_df[(tree_df$label == "SinapisBMAP_alba"), 'label'] <- 'Sinapis_alba_BMAP'
tree_df[(tree_df$label == "ErucaBMAP_vesicaria"), 'label'] <- 'Eruca_vesicaria_BMAP'


for(sp in unique(M$Species)){
  M[M$Species == sp, 'Species'] <- strsplit(sp, '-')[[1]][1]
}
```

Check that there are no mismatches
```{r}
ks_sp <- unique(M$Species)
ks_sp <- data.frame('ks_sp' = ks_sp)
ks_sp$clean <- ks_sp$ks_sp

tr_sp <- unique(tree_df$label)
tr_sp <- tr_sp[!(tr_sp %in% c('', '1.000000'))]
tr_sp <- data.frame('tr_sp' = tr_sp)
tr_sp$clean <- tr_sp$tr_sp

sp_matches <- full_join(ks_sp, tr_sp) 

mask <- (is.na(sp_matches$ks_sp) | is.na(sp_matches$tr_sp))
print(sp_matches[mask, ])
```

# Custom Functions
## Comingling KS
Simulate the distribution prior to species divergence by sampling two or more distributions at given probabilities
```{r}

comingle_KS <- function(
  df = M,
  species = c("Brassica_cretica", "Zilla_macroptera-cdhitest90"),
  weights = c(0.505, 0.5),
  replace = T,
  length_out = 100
  ){
  if(length(species) != length(weights)){
    break
  }
  if (sum(weights) >1){
    print("`weigths` are being normalized to sum to one")
    weights <- (weights/sum(weights))
  }
  sample_each <- length_out*weights
  sample_each <- round(sample_each)
  
  if(sum(sample_each) != length_out){
    print("Sample number changed")
  }
  
  sim_KS <- unlist(purrr::map(seq_along(species), function(i){
    arr <- df[df$Species == species[i], 'KS']
    return(sample(arr, sample_each[i], replace = replace))
  }))

  return(sim_KS)
}
```

### Desmonstrate Functionality 
Demonstrate use (and robustness to incorrectly specified weights)
```{r}
species <- c("Brassica_cretica", "Zilla_macroptera")

test_arr <- comingle_KS(
  df = M,
  species = species,
  weights = c(0.505, 0.5),
  replace = T,
  length_out = 1000)

if(show_demo_plots){
  ggplot(M[M$Species %in% species, ])+
    geom_density(aes(x = KS, fill = Species, color = Species), alpha = 0.5)+
    geom_density(data = data.frame(KS = test_arr
                                   #, Species = rep('Mix', times = length(test_arr))
                                   ),
                 aes(x = KS), color = 'Black')+
    facet_grid(Species~.)+
    scale_fill_manual(values = c('Purple', 'Orange', 'Blue'))+
    scale_color_manual(values = c('Purple', 'Orange', 'Blue'))+
    theme(legend.position = 'bottom') 
}
```

Demonstrate use for an arbitrary number of species (ingroup) 
```{r}
M %>% group_by(Species) %>% tally()

species <- unique(M[M$Group == 'ingroup', 'Species'])

test_arr <- comingle_KS(
  df = M,
  species = species,
  weights = rep(1/length(species), times = length(species)),
  replace = T,
  length_out = 10000)

if(show_demo_plots){
  ggplot(M[M$Species %in% species, ])+
    geom_density(aes(x = KS, fill = Species), alpha = 0.5)+
    geom_density(data = data.frame(KS = test_arr
                                   #, Species = rep('Mix', times = length(test_arr))
                                   ),
                 aes(x = KS), color = '#006400')+
    facet_wrap(Species~.)+
    theme(legend.position = 'none')
}
```

Repeat for outgroup
```{r}
species <- unique(M[M$Group != 'ingroup', 'Species'])


test_arr_outgroup <- comingle_KS(
  df = M,
  species = species,
  weights = rep(1/length(species), times = length(species)),
  replace = T,
  length_out = 10000)

if(show_demo_plots){
  ggplot(M[M$Species %in% species, ])+
    geom_density(aes(x = KS, fill = Species), alpha = 0.5)+
    geom_density(data = data.frame(KS = test_arr
                                   #, Species = rep('Mix', times = length(test_arr))
                                   ),
                 aes(x = KS), color = '#A020F0')+
    facet_wrap(Species~.)+
    theme(legend.position = '')
}
```

Demonstrate capability of resampling distribuiton 
```{r}
if(show_demo_plots){
  # NOTE! to showcase variability I've decreased the sample size
  resample_list <- purrr::map(1:100, function(i){
    comingle_KS(
      df = M,
      species = species,
      weights = rep(1/length(species), times = length(species)),
      replace = T,
      length_out = 100)  
  })
  
  do.call(rbind, purrr::map(seq_along(resample_list), function(i){
    out = data.frame(KS = resample_list[[i]])
    out[['Rep']] = as.character(i)
    return(out)
  })) %>% ggplot(aes(x = KS, group = Rep
                   ))+
    geom_density()
}
```





## Functions to Prepare KSs for ECDF and plot the prepared data
```{r}
# This is adapted from a function I wrote for my thesis. 
# See https://github.com/danielkick/mRNA24hLC
prep_ecdf_ks <- function(
  df = rbind(temp1, temp2),
  data.col = "Corr",
  group.col = "group",
  group1 = "Baseline",   
  group2 = "Compensated",
  colors = c("#006400", 
             #"#67a9cf", 
             "#A020F0"),
  downsample_dist = TRUE) {
  
  # Adapted from:
  # https://rpubs.com/mharris/KSplot
  df <- filter(df, df[[group.col]] %in% c(group1, group2))
  
  data1 <- unlist(df[df[[group.col]] == group1, data.col])
  data2 <- unlist(df[df[[group.col]] == group2, data.col])
  
  # force to be the same length
  if(downsample_dist == TRUE){
    print('Downsampling Distribution')
    min_length = min(length(data1),   length(data2))
    
    if(length(data1) - min_length > 0){
      downsample_idxs <- round(seq(1, length(data1), length.out = min_length))
      data1 <- sort(data1)
      data1 <- data1[downsample_idxs]      
    }
    
    if(length(data2) - min_length > 0){
      downsample_idxs <- round(seq(1, length(data2), length.out = min_length))
      data2 <- sort(data2)
      data2 <- data2[downsample_idxs]      
    }    
  }
  
  ecdf1 <- ecdf(data1)
  ecdf2 <- ecdf(data2)
  
  # used to get the most extreme difference between the two samples
  MostExtremeDiff <- seq(min(data1, data2, na.rm = T), max(data1, data2, na.rm = T), length.out = length(data1))
  x0 <- MostExtremeDiff[which(abs(ecdf1(MostExtremeDiff) - ecdf2(MostExtremeDiff)) == 
                                max(abs(ecdf1(MostExtremeDiff) - ecdf2(MostExtremeDiff))))]
  y0 <- ecdf1(x0)
  y1 <- ecdf2(x0)
  
  # Note that this does not depend on each array being the same length to work.
  graph.df <- rbind(
    data.frame(
    Condition = rep("data1", length(data1)),
    Value = data1
  ),
  data.frame(
    Condition = rep("data2", length(data2)),
    Value = data2
  )
  )
  graph.df[graph.df$Condition == "data1", "Condition"] <- group1
  graph.df[graph.df$Condition == "data2", "Condition"] <- group2
  
  # Run two sided KS test on data
  test.res <- ks.test(data1, data2)
  
  out <- list(
            'group1' = group1,
            'group2' = group2,
            'test.res' = test.res,
            'graph.df' = graph.df,
            'x0' = x0,
            'y0' = y0,
            'y1' = y1,
            'colors' = colors
              )
  return(out)
}

# The fast option will speed up plotting by limiting the number of observations 
plot_ecdf_ks_prep <- function(prep, fast = TRUE, obs_limit = 1000){
  group1 = prep$group1
  group2 = prep$group2
  test.res = prep$test.res
  graph.df = prep$graph.df
  x0 = prep$x0
  y0 = prep$y0
  x1 = prep$x1
  y1 = prep$y1
  colors = prep$colors
  
  if(fast){
    mask <- round(seq(1, nrow(graph.df), length.out = obs_limit))
    graph.df <- graph.df[mask, ]
    }
  
  plt <- 
    ggplot(graph.df)+
    geom_segment(aes(x = x0[1], y = y0[1], xend = x0[1], yend = y1[1]),
                 linetype = "dashed", color = "black", size = 1)+
    geom_point(aes(x = x0[1] , y= y0[1]), color="black", size=2) +
    geom_point(aes(x = x0[1] , y= y1[1]), color="black", size=2) +
    stat_ecdf(aes(x = Value, group = Condition, color = Condition))+
    labs(x = "Sample", 
         y = "ECDF", 
         title = paste("K-S Test", as.character(group1), "vs", as.character(group2), 
                       "\np-value:", as.character(test.res$p.value, digits = 4)))+
    theme_minimal()+
    theme(legend.position = "bottom")+
    scale_color_manual(values = colors)#+
    # theme(text=element_text(family="Calibri Light", size=14)) 
  
  return(plt)
}
```

### Demonstration of Functionality
```{r}
tmp_prep <- prep_ecdf_ks(
  df = M,
  data.col = "KS",
  group.col = "Species",
  group1 = "Brassica_cretica",
  group2 =  "Brassica_hilarionis",
  colors = c("#000099",
             "#990000"), 
  downsample_dist = TRUE)

if(show_demo_plots){
  plot_ecdf_ks_prep(prep = tmp_prep)
  tmp_prep$test.res
}
```

Demonstrate use in conjunction with comingled data
```{r}
df_demo_comingle <- data.frame(
    KS = c(test_arr, test_arr_outgroup),
    Species = c(rep('ingroup', times = length(test_arr)),
                rep('outgroup', times = length(test_arr_outgroup)))
    
  )

df_demo_comingle %>% 
  ggplot(aes(x = KS, fill = Species))+
  geom_density(alpha = 0.5)+
  scale_fill_manual(values = c('#006400', '#A020F0', 'Blue'))+
  theme(legend.position = 'bottom')

tmp_prep <- prep_ecdf_ks(
  df = df_demo_comingle,
  data.col = "KS",
  group.col = "Species",
  group1 =  "ingroup",
  group2 =  "outgroup",
  colors = c("#006400",
             "#A020F0"), 
  downsample_dist = TRUE)

if(show_demo_plots){
  plot_ecdf_ks_prep(prep = tmp_prep)
  tmp_prep$test.res
}
```


```{r}
test_arr_diff <- (sort(test_arr) - sort(test_arr_outgroup))
  
# Compute HDI and ETI
ci_hdi <- ci(test_arr_diff, method = "HDI")
ci_eti <- ci(test_arr_diff, method = "ETI")

if(show_demo_plots){
  data.frame(KS = test_arr_diff) %>% 
    ggplot()+
    geom_vline(xintercept =  0, color = 'blue', linetype = 'dashed')+
    geom_vline(xintercept =  ci_hdi$CI_low,  color = 'firebrick')+
    geom_vline(xintercept =  ci_hdi$CI_high, color = 'firebrick')+
    geom_density(aes(x = KS))+
    coord_cartesian(xlim = c(-0.25, 0.25))
}
```


## Analysis
Prepare pairwise ECDFs of each species
```{r}
sp_comb <- combn(unique(M$Species), 2)
sp_comb <- data.frame(t(sp_comb))
names(sp_comb) <- c("Species_1", "Species_2")

if (file.exists("./sp_comb_list.Rdata")){
  sp_comb_list <- readRDS("./sp_comb_list.Rdata")
} else {
  tictoc::tic()
    # 362.28 sec elapsed (6 minutes)
    sp_comb_list <- map(seq(1,
                            # 100
                            nrow(sp_comb)
                            ), 
      function(i){
        out <- prep_ecdf_ks(
          df = M,
          data.col = "KS",
          group.col = "Species",
          group1 = sp_comb[i, c("Species_1")],
          group2 =  sp_comb[i, c("Species_2")],
          colors = c("#000099", 
                     "#990000"))
        return(out)
    })
  
  tictoc::toc()  

  saveRDS(sp_comb_list, "./sp_comb_list.Rdata")
}
```

```{r}
# merge in ingroup/outgroup labels
M_labs <- M %>% select(-KS) %>% distinct()
sp_comb <-
  left_join(sp_comb, 
            rename(M_labs, 
                   Species_1 = Species,
                   Group_1 = Group)) %>% 
  left_join(rename(M_labs, 
                   Species_2 = Species,
                   Group_2 = Group))
```

# Custom Functions for the Phylogenetic Tree
Find path between root and leaf
```{r}
# given an input species find it's path to the root
leaf_to_root  <- function(
  df = tree_df,
  leaf = 'Brassica_tournefortii'){
  # set up maximum length vectors (clipped later)
  # (should not matter here but better practice than growing them)
  path_branch = rep(NaN, nrow(df))
  path_parent = rep(NaN, nrow(df))
  path_node = rep(NaN, nrow(df))
  
  # define starting node
  i_node = unlist(df[df$label == leaf, ])["node"]
  
  # allow for every branch to be traversed (will not be)
  for(i in seq(1, nrow(df))){
    # print(i)
    j_branch = unlist(df[df$node == i_node, 'branch.length'])
    j_parent = unlist(df[df$node == i_node, 'parent'])
    j_node = unlist(df[df$node == i_node, 'node'])
    
    
    # check if 'node' == 'parent'
    if (i_node == j_parent){
      break
    } else {
      # print(unlist(df[df$node == j_node, ]))   
      
      path_branch[i] = j_branch
      path_parent[i] = j_parent
      path_node[i]   = j_node
      
      i_node <- j_parent
    }
  }
  
  # trim unused indices in vectors
  path_branch = path_branch[1:i-1]
  path_parent = path_parent[1:i-1]
  path_node = path_node[1:i-1]
  
  out = data.frame(
    'from' = path_node,
    'to' = path_parent,
    'branch.length' = path_branch
  )
  return(out)
}
```

Find the weights for all species in a tree to simulate the distribution at a node
```{r}
get_sp_sample_probs <- function(target_node = 74){
  # sweep over the path lists for all species. If the target node is a parent of 
  # the species use `pr_df` to look up the probs for sampling at each branch.
  # return the product of all the probabilities for each species
  
  # go over all species looking for a match
  pr_species_sample <- map(sp_path_list, function(ith_sp_path){
    # check if there is a connection to the target node.
    # if not return 0
    if (!(TRUE %in% (ith_sp_path$to == target_node))){
      ith_sp_pr <- 0
    } else {
      # get all connections up to the one pointing to the target node
      ith_sp_path = ith_sp_path[seq(1, which(ith_sp_path$to == target_node)), ]
      
      #go over all rows in ith_sp_path
      ith_sp_pr <- map(seq(1, nrow(ith_sp_path)), 
                       function(j){
        # get the pr to add to the list
        jth_pr <- unlist(
          pr_df[pr_df$parent == ith_sp_path[j, "to"], 
                as.character(ith_sp_path[j, "from"])]
          )  
      }) %>% 
        unlist()
      # the last item of the cumulative product is the product of the array
      ith_sp_pr <- cumprod(ith_sp_pr)[length(ith_sp_pr)]
    }
    return(as.numeric(ith_sp_pr))
  }) %>% 
    unlist()
  
  return(pr_species_sample)
}
```


# Calculating Species Mixes at Each Node

```{r}
tree_df[is.na(tree_df$branch.length), 'branch.length'] <- 1
# tree_df
```


```{r}
# reverse tree so you can go from parent node to child nodes

parent = 75
df = tree_df

pr_df <- df[, c("parent", "node", "branch.length")] %>% 
  pivot_wider(names_from = "node", values_from = "branch.length")

# normalize lengths to percents
pr_df_sums <- unlist(
  map(seq(1, nrow(pr_df)), function(i){
    pr_df[i, ] %>% select(-parent) %>% sum(na.rm=TRUE)
  })
)

br.len_cols = names(pr_df)[names(pr_df) != 'parent']
pr_df[, br.len_cols] <- pr_df[, br.len_cols]/pr_df_sums

sp_list <- unique(M$Species)
sp_list <- sp_list[!(sp_list %in% c("", "1.000000"))]

sp_path_list <- map(
  sp_list, 
  function(e){
    leaf_to_root(df = tree_df, leaf = e)
    }
  )
```


```{r}
# compute percents for each branch point
parent_nodes <- sort(unique(tree_df$parent))

pr_at_each_branch_point <- do.call(
  rbind, 
  map(
    parent_nodes, 
    function(parent){
      get_sp_sample_probs(target_node = parent)
      }
    )
  )

pr_at_each_branch_point <- data.frame(pr_at_each_branch_point)

names(pr_at_each_branch_point) <- unlist(sp_list)
pr_at_each_branch_point$parent_nodes <- parent_nodes


```

Testing if sample prob calculations worked...
```{r}
# Brassica_tournefortii 1--75--74--
# Erucastrum_virgatum   2-/    |
# Raphanus_raphanistrum 3------/

tmp <- pr_at_each_branch_point %>% 
  filter(parent_nodes == 75)

tmp[, tmp>0]


tmp <- pr_at_each_branch_point %>% 
  filter(parent_nodes == 74)

tmp[, tmp>0]

```


```{r}
# find the last shared ingroup node
leaf_to_root(df = tree_df, leaf = "Arabidopsis_thaliana")

leaf_to_root(df = tree_df, leaf = "Orychrophragmus_violaceae")

leaf_to_root(df = tree_df, leaf = "Brassica_cretica")

# I think 62 is the right node. It should have all the outgroup species
pr_at_each_branch_point %>% 
  filter(parent_nodes == 62) %>% t()
```


## Simulate Ingroup Distribution
Getting a sample with only the ingroup
```{r}
ingroup_sp <- M %>% 
  filter(Group == "ingroup") %>% 
  select(Species) %>% 
  distinct() %>% 
  unlist()


outgroup_sp <- M %>% 
  filter(Group == "outgroup") %>% 
  select(Species) %>% 
  distinct() %>% 
  unlist()

outgroup_mask = (rowSums(pr_at_each_branch_point[, outgroup_sp]) > 0)

# highest_parent_node <- pr_at_each_branch_point[!outgroup_mask, 'parent_nodes'] %>% 
#   max()
# write.csv(pr_at_each_branch_point, "./pr_at_each_branch_point.csv")

# exclude rows with outgroup values
temp <- pr_at_each_branch_point[!outgroup_mask, ingroup_sp]
temp <- temp %>% t()

# which node contains a non-zero percent for all the ingroup species (but none 
# of the outgroup species)
which_node <- map(seq(1, ncol(temp)), function(i){
  if(F %in% temp[, i] > 0){
    # do noting; ith node is not the target node
  }else{
    return(i)  
  }
}) %>% unlist()

ingroup_pr_array <- t(temp)[which_node, ]
```


```{r}
# repeat for outgroup pr_array

temp <- pr_at_each_branch_point[outgroup_mask, outgroup_sp]
temp <- temp %>% t()

# which node contains a non-zero percent for all the ingroup species (but none 
# of the outgroup species)
which_node <- map(seq(1, ncol(temp)), function(i){
  if(F %in% temp[, i] > 0){
    # do noting; ith node is not the target node
  }else{
    return(i)  
  }
}) %>% unlist()

outgroup_pr_array <- t(temp)[which_node, ]

```


## Use ingroup pr array to simulate ancestral distribution
```{r}
sim_length <- M %>% filter(Group == 'ingroup') %>% group_by(Species) %>% tally() %>% select(n)
sim_length <- round(mean(sim_length$n))

sim_ingroup_dist <- comingle_KS(
  df = M,
  species = names(ingroup_pr_array),
  weights = ingroup_pr_array,
  replace = T,
  length_out = sim_length)

if(show_demo_plots){
  ggplot()+geom_density(aes(x = sim_ingroup_dist))
}
```

Resample
```{r}
# simulate a population of distributions
if(show_demo_plots){
  n_sims = 10
  
  sim_ingroup_dist_resampled <- do.call(
    rbind, 
    map(seq(n_sims), function(i){
      out <- data.frame(KS = comingle_KS(
        df = M,
        species = names(ingroup_pr_array),
        weights = ingroup_pr_array,
        replace = T,
        length_out = sim_length)
        )
      out$i <- i
      return(out)
      })
    )
  
  ggplot(sim_ingroup_dist_resampled, aes(x = KS, group = i))+
    geom_density(size = 0.5)
}
```





```{r}
# with Orychrophragmus_violaceae

outgroup_sim_length <- M %>% filter(Group == 'outgroup') %>% group_by(Species) %>% tally() %>% select(n)
outgroup_sim_length <- round(mean(sim_length))


sim_outgroup_dist <- comingle_KS(
  df = M,
  species = names(outgroup_pr_array),
  weights = outgroup_pr_array,
  replace = T,
  length_out = outgroup_sim_length)

if(show_demo_plots){
  ggplot()+geom_density(aes(x = sim_outgroup_dist))
}
```



```{r}
# WITHOUT Orychrophragmus_violaceae

outgroup_sim_length_wo_ory <- M %>% 
  filter(Group == 'outgroup') %>% 
  filter(Species != 'Orychrophragmus_violaceae') %>% 
  group_by(Species) %>% tally() %>% select(n)
outgroup_sim_length_wo_ory <- round(mean(sim_length))


outgroup_pr_array_wo_ory <- outgroup_pr_array[names(outgroup_pr_array) != 'Orychrophragmus_violaceae']

sim_outgroup_dist_wo_ory <- comingle_KS(
  df = M,
  species = names(outgroup_pr_array_wo_ory),
  weights = outgroup_pr_array_wo_ory,
  replace = T,
  length_out = outgroup_sim_length_wo_ory)

if(show_demo_plots){
  ggplot()+geom_density(aes(x = sim_outgroup_dist_wo_ory))
}
```

```{r}
# Plot all three sim dist for comparison

if(show_demo_plots){
  ggplot()+
    geom_density(aes(x = sim_ingroup_dist), color = '#006400')+
    geom_density(aes(x = sim_outgroup_dist), color = '#A020F0')+
    geom_density(aes(x = sim_outgroup_dist_wo_ory), color = '#000000')
}
# don't do outgroup wo ory
```


# Ingroup vs Outgroup
```{r}
if(show_demo_plots){
  ggplot()+
    geom_density(aes(x = sim_ingroup_dist), color = '#006400')+
    geom_density(aes(x = sim_outgroup_dist), color = '#A020F0')
}
```

```{r}
temp <- data.frame(
  KS = c(sim_ingroup_dist, sim_outgroup_dist),
  Dist = c(rep("ingroup", times = length(sim_ingroup_dist)),
           rep("outgroup", times = length(sim_outgroup_dist))
                    ))


tmp_prep <- prep_ecdf_ks(
  df = temp,
  data.col = "KS",
  group.col = "Dist",
  group1 = "ingroup",
  group2 =  "outgroup",
  colors = c("#006400",
             "#A020F0"), 
  downsample_dist = TRUE)

plt <- plot_ecdf_ks_prep(prep = tmp_prep)
plt
tmp_prep$test.res
ggsave("./output/ingroup_vs_outgroup.svg", plt, width = 11, height = 8.5)
```


```{r}
# Ingroup vs ory

temp2 <- M %>% select(-Group, -Subtribe) %>% filter(Species == 'Orychrophragmus_violaceae') %>% rename(Dist = Species)

temp2 <- rbind(temp, temp2)



# vs ingroup

# tmp_prep <- prep_ecdf_ks(
#   df = temp2,
#   data.col = "KS",
#   group.col = "Dist",
#   group1 = "Orychrophragmus_violaceae",
#   group2 =  "ingroup",
#   colors = c("#000000",
#              "#006400"), 
#   downsample_dist = TRUE)
# 
# tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)

tmp_prep$test.res


tmp_plt <- tmp_plt+
  labs(title = 'K-S Test *Orychrophragmus violaceae* vs Ingroup',
       subtitle = 'p <2.2e-16')+
  scale_color_manual(
    values =  c("#006400", "#000000"),
    limits = c("ingroup", "Orychrophragmus_violaceae"),
    labels = c("Ingroup", "*Orychrophragmus violaceae*")
  )+
  md_theme_minimal()+
  theme(legend.position = 'bottom')


tmp_plt
ggsave("./output/Orychrophragmus_violaceae_vs_ingroup.svg", tmp_plt, width = 11, height = 8.5)
# ggsave(tmp_plt, './output/KS_Contrast_Oviolaceae_vs_ingroup.pdf')


```


```{r}
# vs outgroup

# tmp_prep <- prep_ecdf_ks(
#   df = temp2,
#   data.col = "KS",
#   group.col = "Dist",
#   group1 = "Orychrophragmus_violaceae",
#   group2 =  "outgroup",
#   colors = c("#000000",
#              "#A020F0"
#              ), 
#   downsample_dist = TRUE)
# 
# 
# tmp_prep$test.res
# 
# tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)




tmp_plt <- tmp_plt+
  labs(title = 'K-S Test *Orychrophragmus violaceae* vs Outgroup',
       subtitle = 'p <2.2e-16')+
  scale_color_manual(
    values =  c("#A020F0", "#000000"),
    limits = c("outgroup", "Orychrophragmus_violaceae"),
    labels = c("Outgroup", "*Orychrophragmus violaceae*")
  )+
  md_theme_minimal()+
  theme(legend.position = 'bottom')

tmp_plt
ggsave("./output/Orychrophragmus_violaceae_vs_outgroup.svg", tmp_plt, width = 11, height = 8.5)
```


# Contrast Subtribes

```{r}
# setup utility functions
get_subtribe_pr_array <- function(subtribe_name = "Cakile"){
  # find all the species in a given subtribe
  ingroup_sp <- M %>% 
    filter(Subtribe == subtribe_name) %>% 
    select(Species) %>% 
    distinct() %>% 
    unlist()
  # check if the mix is actually just a single species (Subtribe Psychine)
  if(length(ingroup_sp) == 1){
    # If there's only a single species
    ingroup_pr_array <- 1.00
    names(ingroup_pr_array) <- ingroup_sp
  }else{
    # then define the outgroup as all those not in the ingroup
    outgroup_sp <- M %>% 
      filter(!(Species  %in% ingroup_sp)) %>% 
      select(Species) %>% 
      distinct() %>% 
      unlist()
    
    outgroup_mask = (rowSums(pr_at_each_branch_point[, outgroup_sp]) > 0)
    
    # exclude rows with outgroup values
    temp <- pr_at_each_branch_point[!outgroup_mask, ingroup_sp]
    temp <- temp %>% t()
    
    # which node contains a non-zero percent for all the ingroup species (but none 
    # of the outgroup species)
    which_node <- map(seq(1, ncol(temp)), function(i){
      if(F %in% temp[, i] > 0){
        # do noting; ith node is not the target node
      }else{
        return(i)  
      }
    }) %>% unlist()
    
    ingroup_pr_array <- t(temp)[which_node, ]
    

  }
  return(ingroup_pr_array)  
}

get_subtribe_sim_length <- function(subtribe_name = "Cakile"){
  sim_length <- M %>% filter(Subtribe == subtribe_name) %>% group_by(Species) %>% tally() %>% select(n)
  sim_length <- round(mean(sim_length$n))
  return(sim_length)
}

get_subtribe_sim_dist <- function(subtribe_pr_array, sim_length){
  ingroup_sp_names <- names(ingroup_pr_array)
  
  # if there's only one species in this set return in from M
  if(length(ingroup_sp_names) == 1){
    sim_ingroup_dist <- M %>% 
      filter(Species == ingroup_sp_names) %>% 
      select(KS) %>% 
      unlist() %>% 
      sort()
  }else{
    sim_ingroup_dist <- comingle_KS(
      df = M,
      species = ingroup_sp_names,
      weights = ingroup_pr_array,
      replace = T,
      length_out = sim_length)    
  }
  return(sim_ingroup_dist) 
}

wrapper_subtribe_sim_dist <- function(subtribe_name = "Cakile"){
  subtribe_pr_array <- get_subtribe_pr_array(  subtribe_name = subtribe_name)
  sim_length        <- get_subtribe_sim_length(subtribe_name = subtribe_name)
  
  output_dist       <- get_subtribe_sim_dist(subtribe_pr_array = subtribe_pr_array, 
                                             sim_length = sim_length)  
  return(output_dist)
}


# Get dist for all subtribes
needed_subtribes <- c("Brassica", "Cakile", "Crambe", "Savignya", "Zilla", "Psychine", "Vella")

needed_subtribe_dists <- map(
  needed_subtribes, 
  function(ith_subtribe){
    # print(ith_subtribe)
    out = wrapper_subtribe_sim_dist(subtribe_name = ith_subtribe)
    return(out)
    })

names(needed_subtribe_dists) <- needed_subtribes

# Set up all comparisons 
subtribe_comparisons <- rbind(
  data.frame(
    Var1 = c("Vella", "Zilla", "Psychine", "Savignya", "Cakile", "Crambe"),
    Var2 = c("Zilla", "Psychine", "Savignya", "Cakile", "Crambe", "Brassica")
  ),
  expand.grid(
  "Brassica", 
  c("Cakile", "Crambe", "Savignya", "Zilla", "Psychine", "Vella")
  )
)
  
needed_subtribe_plts <- map(
  seq(1, nrow(subtribe_comparisons)), 
  function(i){
      
    Group1 <- subtribe_comparisons[i, "Var1"]
    Group2 <- subtribe_comparisons[i, "Var2"]
    
    KS1 <- needed_subtribe_dists[[Group1]]
    KS2 <- needed_subtribe_dists[[Group2]]
    
    temp <- data.frame(
      Dist = c(rep(Group1, times = length(KS1)),
               rep(Group2, times = length(KS2))),
      KS   = c(KS1, KS2)
    )
    
    
    tmp_prep <- prep_ecdf_ks(
      df = temp,
      data.col = "KS",
      group.col = "Dist",
      group1 = Group1,
      group2 = Group2,
      colors = c("#000000",
                 "#990000"
                 ), 
      downsample_dist = TRUE)
    
    tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)
    
    tmp_plt <- tmp_plt+
      labs(title = paste0('*',Group1,'* vs *', Group2, '*'),
           subtitle = paste0('p = ', as.character(tmp_prep$test.res$p.value)))+
      scale_color_manual(
        values =  c("#000000", "#990000"),
        limits = c(Group1, Group2),
        labels = c(paste0('*', Group1, '*'), paste0('*', Group2, '*'))
      )+
      md_theme_minimal()+
      theme(legend.position = 'bottom')
    
    return(tmp_plt)
})

plt <- cowplot::plot_grid(plotlist = needed_subtribe_plts)
ggsave("./output/subtribe_vs_subtribe.svg", plt, width = 11, height = 8.5)



  

extra_subtribe_plts_vs_ingroup <- map(
  seq(1, length(needed_subtribe_dists)), 
  function(i){
      
    Group1 <- names(needed_subtribe_dists)[i]
    Group2 <- "ingroup"
    
    KS1 <- needed_subtribe_dists[[Group1]]
    KS2 <- sim_ingroup_dist
    
    temp <- data.frame(
      Dist = c(rep(Group1, times = length(KS1)),
               rep(Group2, times = length(KS2))),
      KS   = c(KS1, KS2)
    )
    
    
    tmp_prep <- prep_ecdf_ks(
      df = temp,
      data.col = "KS",
      group.col = "Dist",
      group1 = Group1,
      group2 = Group2,
      colors = c("#000000",
                 "#990000"
                 ), 
      downsample_dist = TRUE)
    
    tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)
    
    tmp_plt <- tmp_plt+
      labs(title = paste0('*',Group1,'* vs *', Group2, '*'),
           subtitle = paste0('p = ', as.character(tmp_prep$test.res$p.value)))+
      scale_color_manual(
        values =  c("#000000", "#006400"),
        limits = c(Group1, Group2),
        labels = c(paste0('*', Group1, '*'), paste0('*', Group2, '*'))
      )+
      md_theme_minimal()+
      theme(legend.position = 'bottom')
    
    return(tmp_plt)
})

extra_subtribe_plts_vs_outgroup <- map(
  seq(1, length(needed_subtribe_dists)), 
  function(i){
      
    Group1 <- names(needed_subtribe_dists)[i]
    Group2 <- "outgroup"
    
    KS1 <- needed_subtribe_dists[[Group1]]
    KS2 <- sim_outgroup_dist
    
    temp <- data.frame(
      Dist = c(rep(Group1, times = length(KS1)),
               rep(Group2, times = length(KS2))),
      KS   = c(KS1, KS2)
    )
    
    
    tmp_prep <- prep_ecdf_ks(
      df = temp,
      data.col = "KS",
      group.col = "Dist",
      group1 = Group1,
      group2 = Group2,
      colors = c("#000000",
                 "#990000"
                 ), 
      downsample_dist = TRUE)
    
    tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)
    
    tmp_plt <- tmp_plt+
      labs(title = paste0('*',Group1,'* vs *', Group2, '*'),
           subtitle = paste0('p = ', as.character(tmp_prep$test.res$p.value)))+
      scale_color_manual(
        values =  c("#000000", "#A020F0"),
        limits = c(Group1, Group2),
        labels = c(paste0('*', Group1, '*'), paste0('*', Group2, '*'))
      )+
      md_theme_minimal()+
      theme(legend.position = 'bottom')
    
    return(tmp_plt)
})


plt <- cowplot::plot_grid(plotlist = extra_subtribe_plts_vs_ingroup)
ggsave("./output/subtribe_vs_ingroup.svg", plt, width = 11, height = 8.5)
plt <- cowplot::plot_grid(plotlist = extra_subtribe_plts_vs_outgroup)
ggsave("./output/subtribe_vs_outgroup.svg", plt, width = 11, height = 8.5)


```




Since outgroup doesn't contain any of the ingroup data what would the ancestral distribution look like?
```{r}
# pr array for the root node:
root_pr_array <- pr_at_each_branch_point[1, ] %>% select(-parent_nodes) 

if(mean(root_pr_array > 0) != 1){
  print("Root Node Incorrectly Selected")
} else {
  
  root_pr_array <- unlist(root_pr_array)
  
  sim_length <- M %>% group_by(Species) %>% tally() %>% select(n)
  sim_length <- round(mean(sim_length$n))
  
  sim_ancestral_dist <- comingle_KS(
    df = M,
    species = names(root_pr_array),
    weights = root_pr_array,
    replace = T,
    length_out = sim_length)
  
  
  
  
  extra_subtribe_plts_vs_ancestral <- map(
    seq(1, length(needed_subtribe_dists)), 
    function(i){
        
      Group1 <- names(needed_subtribe_dists)[i]
      Group2 <- "Ancester"
      
      KS1 <- needed_subtribe_dists[[Group1]]
      KS2 <- sim_ancestral_dist
      
      temp <- data.frame(
        Dist = c(rep(Group1, times = length(KS1)),
                 rep(Group2, times = length(KS2))),
        KS   = c(KS1, KS2)
      )
      
      
      tmp_prep <- prep_ecdf_ks(
        df = temp,
        data.col = "KS",
        group.col = "Dist",
        group1 = Group1,
        group2 = Group2,
        colors = c("#000000",
                   "#990000"
                   ), 
        downsample_dist = TRUE)
      
      tmp_plt <- plot_ecdf_ks_prep(prep = tmp_prep)
      
      tmp_plt <- tmp_plt+
        labs(title = paste0('*',Group1,'* vs *', Group2, '*'),
             subtitle = paste0('p = ', as.character(tmp_prep$test.res$p.value)))+
        scale_color_manual(
          values =  c("#000000", "#990000"),
          limits = c(Group1, Group2),
          labels = c(paste0('*', Group1, '*'), paste0('*', Group2, '*'))
        )+
        md_theme_minimal()+
        theme(legend.position = 'bottom')
      
      return(tmp_plt)
  })
  
  plt <- cowplot::plot_grid(plotlist = extra_subtribe_plts_vs_ancestral)
  ggsave("./output/subtribe_vs_ancestral.svg", plt, width = 11, height = 8.5)
}
```


# Record all simulated distributions

```{r}
write_dist <- function(dist_to_write = sim_ingroup_dist,
                       save_name = 'dist_sim_ingroup'){
  out <- data.frame(dist_to_write)
  names(out) <- save_name
  write.csv(out, file = paste0('./output/', save_name, '.csv'), row.names = F)
}

tmp_list_names <- list(
  'sim_ingroup_dist',
  'sim_outgroup_dist',
  'sim_outgroup_dist_wo_ory',
  'sim_ancestral_dist') 
tmp_list <- list(
  sim_ingroup_dist,
  sim_outgroup_dist,
  sim_outgroup_dist_wo_ory,
  sim_ancestral_dist)

for(i in seq(1, length(tmp_list))){
  write_dist(
    dist_to_write = tmp_list[[i]],
    save_name = tmp_list_names[[i]]
    )
}

for(i in seq(1, length(needed_subtribe_dists))){
  write_dist(
    dist_to_write = needed_subtribe_dists[[i]],
    save_name = names(needed_subtribe_dists)[i]
    )
}
```


```{r}
dist_lengths <- unlist(purrr::map(c(
  "Brassica.csv",
  "Cakile.csv",
  "Crambe.csv",
  "dist_sim_ingroup.csv",
  "Psychine.csv",
  "Savignya.csv",
  "sim_ancestral_dist.csv",
  "sim_ingroup_dist.csv",
  "sim_outgroup_dist.csv",
  "sim_outgroup_dist_wo_ory.csv",
  "Vella.csv",
  "Zilla.csv"), function(e){
  tmp = read.csv(paste0("./output/", e))
  return(dim(tmp)[1])
}))

min(dist_lengths)
max(dist_lengths)


```


